---
title: "Analysis of TF and kinases among known oncogenes"
date: "04.02.24"
output: 
  html_document:
  latex_engine : xelatex
df_print: default
highlight: zenburn
toc: FALSE
mainfont : NanumGothic    
editor_options: 
  markdown: 
  wrap: 72
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(VennDiagram)
library(gridExtra)
library(rvest)
library(ggbreak) 
library(patchwork) 
```

```{r setwd, include=FALSE}
main_dir <- dirname(rstudioapi::getSourceEditorContext()$path) 
setwd(main_dir)
```

### Data preparation

```{r data import}
TF <- read.csv("../../Data/Raw_data/Human_TFs.txt", sep="\t")
COSMIC <- read.csv("../../Data/Raw_data/Cancer_Gene_census_COSMIC.tsv", sep = "\t")
oncoKB <- read.csv("../../Data/Raw_data/cancerGeneList_OncoKB.tsv", sep = "\t")
```


### TFs (enrichment among oncogenes vs in all genes)

Total TF:

```{r TF count}
total_TF <- length(unique(TF$HGNC.symbol))
total_TF
```

In Gencode there are 20241 genes. TFs are 13.7%:

```{r}
276500/20241
```
Total oncogenes from COSMIC+oncoKB:

```{r oncogenes count}
COSMIC <- COSMIC[COSMIC$Role.in.Cancer %in% c(c("oncogene, fusion", "oncogene")), ]
COSMIC_oncogenes <- COSMIC$Gene.Symbol
oncoKB_oncogenes <- oncoKB$Hugo.Symbol[oncoKB$Is.Oncogene == "Yes" & oncoKB$Is.Tumor.Suppressor.Gene != "Yes"]
oncogenes_total <- union(COSMIC_oncogenes, oncoKB_oncogenes)
total_oncogenes <- length(oncogenes_total)
total_oncogenes
```

How many oncogenes are TF (40.5%):

```{r oncogenes are TF}
oncogenes_TF <- intersect(oncogenes_total, unique(TF$HGNC.symbol))
TFs_in_oncogenes <- length(oncogenes_TF)
TFs_in_oncogenes
```

Do oncogenes more often appears TFs:

```{r fisher test}
set_1_TF <- 2765
set_1_genes <- 20241 - 2765
set_2_TF <- 181
set_2_oncogenes <- 447 - 181

contingency_table_TF <- matrix(c(set_2_TF, set_1_TF, set_2_oncogenes,
                              set_1_genes),
                            nrow = 2)
colnames(contingency_table_TF) <- c('TF', 'not TF')
rownames(contingency_table_TF) <- c('Oncogenes', 'All genes')

fisher.test(contingency_table_TF)
```


### Kinase (enrichment among oncogenes vs in all genes)

Total kinases:

```{r kinase count}
webpage <- read_html("http://www.kinhub.org/kinases.html")
kinase <- html_table(webpage)[[1]]
kinase <- kinase %>%
  dplyr::select("xName", "HGNC Name", "Group", "Family")
colnames(kinase)[2] <- "HGNC_Name"

# Some kinases have no name in "HGNC Name" column, we take it than from "xName" column (impute missing values)

kinase <- kinase %>%
  mutate(HGNC_Name = ifelse(HGNC_Name == "", xName, HGNC_Name))

# Some of values were not unique due to different columns (with their name) being different in the full data set
# Let make them unique by using other column for naming of these

kinase$Name <- ifelse(duplicated(kinase$HGNC_Name) & !duplicated(kinase$xName), kinase$xName, kinase$HGNC_Name)
kinase <- kinase[, -c(1,2)]

# Count total N of kinase

length(unique(kinase$Name))
```

In Gencode there are 20241 genes. Kinases are 2.6%:

```{r}
53600/20241
```

Total oncogenes from COSMIC+oncoKB - 447.

How many oncogenes are kinases (19%):

```{r oncogenes are kinases}
oncogenes_kinases <- intersect(oncogenes_total, unique(kinase$Name))
kinases_in_oncogenes <- length(oncogenes_kinases)
kinases_in_oncogenes
```


```{r fisher test 2}
set_1_kinases <- 536
set_1_genes <- 20241 - 536
set_2_kinases <- 85
set_2_oncogenes <- 447 - 85

contingency_table_kinases <- matrix(c(set_2_kinases, set_1_kinases, set_2_oncogenes,
                              set_1_genes),
                            nrow = 2)

colnames(contingency_table_kinases) <- c('Kinase', 'not Kinase')
rownames(contingency_table_kinases) <- c('Oncogenes', 'All genes')

fisher.test(contingency_table_kinases)
```

CONCLUSION: Known oncogenes have enrichment by kinases and TFs.

## Visualize

```{r}
contingency_table_TF[1,] <- contingency_table_TF[1,]*100/(contingency_table_TF[1,1]+ contingency_table_TF[1,2])
contingency_table_TF[2,] <- contingency_table_TF[2,]*100/(contingency_table_TF[2,1]+ contingency_table_TF[2,2])

barplot(t(contingency_table_TF), beside = TRUE, col = c("#fed9a5", "#b3cee3"), legend = TRUE, ylab = "Genes (%)") 
```
```{r}
contingency_table_kinases[1,] <- contingency_table_kinases[1,]*100/( contingency_table_kinases[1,1]+ contingency_table_kinases[1,2])
contingency_table_kinases[2,] <- contingency_table_kinases[2,]*100/( contingency_table_kinases[2,1]+ contingency_table_kinases[2,2])

barplot(t(contingency_table_kinases), beside = TRUE, col = c("#fed9a5", "#b3cee3"), legend = TRUE, ylab = "Genes (%)") 
```
