---
title: "Statistics of known oncogenes from COSMIC and oncoKB in our data"
output: 
  html_document:
  latex_engine : xelatex
df_print: default
highlight: zenburn
toc: FALSE
mainfont : NanumGothic    
editor_options: 
  markdown: 
  wrap: 72
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(VennDiagram)
library(gridExtra)
```

```{r setwd, include=FALSE}
main_dir <- dirname(rstudioapi::getSourceEditorContext()$path) 
setwd(main_dir)
```

### Data preparation

Read data and leave only regions with known genes (data has been merged with oncogene  containing datasets COSMIC and oncoKB + region thresholding and interval merging is done)

```{r data import and remove regions without genes}
data <- read.csv("../Data/Processed_data/Table_expression_COSMIC_oncoKB_merged_intervals.csv")
data <- data[data$gene_name!="", ]

cancer_types <- read.csv("../Data/Raw_data/Table_for_cancer_types.csv")
cancer_types <- cancer_types %>% 
  select("name", "cancer_type")
data <- left_join(data, cancer_types, join_by("sample" == "name"))
```

Number os samples per cancer type has been precomputed on our data before. Leave only cancer types where at least 20 high copy regions are present:

```{r cancer type selection}
n_copies_cancer_types <- read.csv("../Data/Raw_data/Number_of_copies_per_sample_for_cancer_type.csv")

summary_samples_cancer <- n_copies_cancer_types %>%
  group_by(Cancer_type) %>%
  summarize(
    samples = sum(Sample_number)
  ) %>%
  ungroup()

threshold_of_N_samples <- 20
cancers_to_keep <- summary_samples_cancer$Cancer_type[summary_samples_cancer$samples >= threshold_of_N_samples]

data <- data[data$cancer_type %in% cancers_to_keep, ]
```

Note: 17 out of 28 cancer types are left after this step (kept: "BLCA" "BRCA" "CESC" "COAD" "ESCA" "GBM"  "HNSC" "LGG"  "LIHC" "LUAD" "LUSC" "OV"   "PAAD" "SARC" "SKCM" "STAD" "UCEC")

### Let's group data by cancer type and unique regions to see how many genes in our data are labeled as oncogenes:
- only in COSMIC 
- only in oncoKB 
- in both data sets
- in none of data sets

```{r count genes per region and cancer type}
data$coordinates <- paste(as.character(data$chr), ":", data$startpos, "-", data$endpos)

count_per_region <- data %>%
  group_by(cancer_type, coordinates) %>%
  summarize(
    in_COSMIC = sum((Tier.COSMIC == "1" | Tier.COSMIC == "2") & OncoKB.Annotated.oncoKB != "Yes"),
    in_oncoKB = sum(OncoKB.Annotated.oncoKB == "Yes" & Tier.COSMIC == "No_data"),
    in_both = sum((Tier.COSMIC == "1" | Tier.COSMIC == "2") & OncoKB.Annotated.oncoKB == "Yes"),
    nowhere = sum(Tier.COSMIC == "No_data" & (OncoKB.Annotated.oncoKB == "No_data" | OncoKB.Annotated.oncoKB == "No")),
  ) %>%
  ungroup()

count_per_region
```

This many regions over all cancers got at least one oncogene:

```{r}
sum(count_per_region$in_COSMIC>0 | count_per_region$in_oncoKB>0 | count_per_region$in_both>0)
```

This many regions over all cancers got NO oncogenes from COSMIC or oncoKB:

```{r}
sum(count_per_region$in_COSMIC==0 & count_per_region$in_oncoKB==0 & count_per_region$in_both==0)
```
Let's see how many it is per each cancer type:

```{r count genes per region and cancer type summary and graph}
summary_count_per_region <- count_per_region %>%
  group_by(cancer_type) %>%
  summarize(
    regions_with_oncogene = sum(in_COSMIC>0 | in_oncoKB>0 | in_both>0),
    regions_without_oncogene = sum(in_COSMIC==0 & in_oncoKB==0 & in_both==0),
  ) %>%
  ungroup()

summary_count_per_region
```

```{r }
summary_count_per_region_long <- pivot_longer(summary_count_per_region, cols = c("regions_with_oncogene", "regions_without_oncogene"), names_to = "presence_of_oncogenes", values_to = "regions")

ggplot(summary_count_per_region_long, aes(fill=presence_of_oncogenes, y=regions, x=cancer_type)) + 
    geom_bar(position=position_fill(reverse = TRUE), stat="identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Acessing distribution of region length and copy number for regions with at least one oncogene and without
Of note: there are 3 regions that are exactly having the same borders within the same cancer type, but they are coming from different samples, their copy number is different. I take therefore mean of copy number. For other region it doesnt make any difference as they have only one sample. 

It is:

GBM 7 : 55225629 - 55798100         has oncogene      22 and 42 copies

STAD 11 : 34666863 - 35709082       no oncogene       25 and 27 copies

UCEC 17 : 37671490 - 37974720       has oncogene      16 and 21 copies 


```{r distribution}
count_per_region$coordinates_2 <- sub(".* : ", "", count_per_region$coordinates)
count_per_region <- separate(count_per_region, coordinates_2, into = c("start", "end"), sep = " - ")
count_per_region$start <- as.integer(count_per_region$start)
count_per_region$end <- as.integer(count_per_region$end)
count_per_region$length <- count_per_region$end - count_per_region$start
count_per_region <- count_per_region %>%
  mutate(has_oncogene = ifelse(in_COSMIC > 0 | in_oncoKB > 0 | in_both > 0, "Yes", "No"))
count_per_region$ID <- paste(count_per_region$cancer_type, count_per_region$coordinates)
data$ID <- paste(data$cancer_type, data$coordinates)

data_check <- data %>%
  group_by(ID) %>%
  summarize(
  nUniqueMajor = n_distinct(nMajor),
  copy_number = as.integer(mean(nMajor))
  ) %>%
  ungroup()

count_per_region <- left_join(count_per_region, data_check, by = "ID")

ggplot(count_per_region, aes(x=length, color=has_oncogene)) + 
  geom_histogram(binwidth=20000) +
  ggtitle("Distributions of length of regions with and without known oncogene")
```
```{r distribution 2}
ggplot(count_per_region, aes(x=log10(copy_number), color=has_oncogene, fill=has_oncogene))+
  geom_density(alpha=0.4) +
  ggtitle("Distributions of copy number of regions with and without known oncogene")
```

Remember: there are 3800 regions without oncogene and 2253 with oncogene.

### Region length vs copy number


```{r}
ggplot(count_per_region, aes(x=log2(length), y=log2(copy_number), color=has_oncogene)) + 
    geom_point(size=1) 
```

### Distribution of number of oncogene per region

```{r distribution 3}
count_per_region <- count_per_region %>%
  mutate(number_of_oncogenes = count_per_region$in_COSMIC+count_per_region$in_oncoKB+count_per_region$in_both)

ggplot(count_per_region, aes(x=number_of_oncogenes)) + 
  geom_histogram(binwidth=1) +
  ggtitle("Distributions of number of oncogenes per region")
```

### Distribution of number of oncogene per sample

```{r distribution 4}
count_per_sample <- data %>%
  group_by(cancer_type, sample) %>%
  summarize(
    in_COSMIC = sum((Tier.COSMIC == "1" | Tier.COSMIC == "2") & OncoKB.Annotated.oncoKB != "Yes"),
    in_oncoKB = sum(OncoKB.Annotated.oncoKB == "Yes" & Tier.COSMIC == "No_data"),
    in_both = sum((Tier.COSMIC == "1" | Tier.COSMIC == "2") & OncoKB.Annotated.oncoKB == "Yes"),
    nowhere = sum(Tier.COSMIC == "No_data" & (OncoKB.Annotated.oncoKB == "No_data" | OncoKB.Annotated.oncoKB == "No")),
  ) %>%
  ungroup()

count_per_sample <- count_per_sample %>%
  mutate(number_of_oncogenes = count_per_sample$in_COSMIC+count_per_sample$in_oncoKB+count_per_sample$in_both)

ggplot(count_per_sample, aes(x=number_of_oncogenes)) + 
  geom_histogram(binwidth=1) +
  ggtitle("Distributions of number of oncogenes per sample")
```

All the exploratory statistic above was base on the oncogens present at least in one of the data bases. 

### If there are multiple regions are they likely to have one or many oncogenes

```{r }
data$coordinates <- sub(".*: ", "", data$ID)

cancer_sample_summary <- data %>%
  group_by(cancer_type, sample, coordinates) %>%
  summarize(
    in_COSMIC = sum((Tier.COSMIC == "1" | Tier.COSMIC == "2") & OncoKB.Annotated.oncoKB != "Yes"),
    in_oncoKB = sum(OncoKB.Annotated.oncoKB == "Yes" & Tier.COSMIC == "No_data"),
    in_both = sum((Tier.COSMIC == "1" | Tier.COSMIC == "2") & OncoKB.Annotated.oncoKB == "Yes"),
    nowhere = sum(Tier.COSMIC == "No_data" & (OncoKB.Annotated.oncoKB == "No_data" | OncoKB.Annotated.oncoKB == "No")),
    regions_count = n(),
  ) %>%
  ungroup()

cancer_sample_summary <- cancer_sample_summary %>%
  mutate(oncogenes_total = in_COSMIC + in_oncoKB + in_both)

cor.test(cancer_sample_summary$regions_count, cancer_sample_summary$oncogenes_total)
```

There seems to be weak correlation between number of regions in the sample and number of oncogenes. Let's see the graph:

```{r }
ggplot(cancer_sample_summary, aes(x = regions_count)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Regions Count",
       x = "Regions per Sample Count",
       y = "Frequency")
```

```{r }
ggplot(cancer_sample_summary, aes(x = oncogenes_total)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Oncogenes Count",
       x = "Oncogenes per Sample Count",
       y = "Frequency")
```
```{r}
my_colors <- c(
  "#1f78b4", "#33a02c", "#e31a1c", "#ff7f00", "#6a3d9a",
  "#a6cee3", "#b2df8a", "#fb9a99", "#fdbf6f", "#cab2d6",
  "#008080", "#d73027", "#4575b4", "#91cf60", "#fee08b",
  "#313695", "#dd1c77"
)

ggplot(cancer_sample_summary, aes(x=regions_count, y=oncogenes_total, color=cancer_type)) + 
    geom_point(size=2) +
    scale_color_manual(values = my_colors) +
    ggtitle("Number of oncogenes per sample vs numer of regions per sample")  
```

